---
title: "Chromatin bound protein model simulated data"
author: "Sam Haynes"
date: "2023-03-15"
output: html_document
---

# Overview

The overall picture is that, we have chromatin-enriched samples MS-level protein intensities. But because each sample has been enriched to a different degree, some abundance reflect both the behavior of that protein (more or less abundant in a cell line) but also the technical preparation of that cell line ( more or less  chromatin-enriched).
 
So the whole attempt of the model is to determine how well each sample is prepared (chromatin-enriched) to then figure out if a protein in more abundant in that cell line. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(here)

set.seed(324)


real_protein_dataset <- readRDS(here("data/input/real_data_list.rds"))

```


```{r output-observed-data}
# Real data set is imbalanced. Recommend randomly selecting 2 of the 7 samples for cell line 7. Don't lose too much data and avoids over emphasis on protein behaviour in this cell line. Alternatives require more complex statistics! 

tibble(cell_line = real_protein_dataset$cell_line) |> 
  group_by(cell_line) |> 
  summarise(count = n()) |>
  arrange(desc(count))

resampled_real_dataset <- tibble(sample = rep(1:115, 50),
       cell_line = rep(real_protein_dataset$cell_line, 50),
       input_enrichment = rep(real_protein_dataset$input_enrichment, 50),
       protein = rep(1:50, each = 115),
       prot_intensity = c(real_protein_dataset$prot_intensity)) |>
  group_by(cell_line) |>
  filter(sample %in% sample(unique(sample), 2)) |>
  arrange(protein, cell_line) |>
  ungroup()

real_protein_dataset$N <- 110

real_protein_dataset$prot_intensity <- resampled_real_dataset$prot_intensity |>
  matrix(nrow = 110)

real_protein_dataset$input_enrichment <- resampled_real_dataset$input_enrichment[1:110]

real_protein_dataset$cell_line <- resampled_real_dataset$cell_line[1:110]

saveRDS(real_protein_dataset, here("data/output/subsampled_real_data_list.rds"))
```

```{r ingest_all_raw_data}
full_raw_data <- read_csv(here("data/input/input_report.csv"))

perfect_data_set <- full_raw_data |>
  pivot_longer(-ProteinGroup, names_to = "sample", values_to = "prot_intensity") |>
  separate(sample, sep ="_", into = c("tissue_name", "sample"), extra="merge") |>
  separate(sample, sep ="_(?=[1-9]$)", into = c("cell_line_name", "rep_name")) |>
  group_by(ProteinGroup) |>
  filter(all(!is.na(prot_intensity)),
         !str_detect(tissue_name, "pool")) |>
  mutate(tissue_name = factor(tissue_name),
         cell_line_name = factor(cell_line_name),
         ProteinGroup = factor(ProteinGroup)) |>
  ungroup()

prepped_raw_data_set <- perfect_data_set |>
  mutate(tissue = unclass(tissue_name),
         cell_line = unclass(cell_line_name),
         protein_id = unclass(ProteinGroup)) |>
  group_by(ProteinGroup,cell_line, tissue) |>
  mutate(rep = 1:length(rep_name)) |>
  ungroup()
  
```

```{r output, eval=FALSE}
saveRDS(prepped_raw_data_set, here("data/output/perfect_real_data_set.rds"))
```
