---
title: "Chromatin Bound Proteins Model Update"
output: pdf_document
date: "2023-08-07"
---

## Overview

The Bayesian model deduces three variables:

- $r_p$ : The cytoplasmic to nuclear ratio of a protein
- $\epsilon_s$ : The chromatin enrichment of each sample
- $I^{chr}_{pc}$ : The cell line specific intensity of a proteinin the chromatin

$$I^{Tot}_{ps}=\epsilon_sI^{chr}_{pc}+(1-\epsilon_s)r_pI^{chr}_{pc}$$

Intensities are normalised using a DESeq2-like median of medians before being placed inside the model.

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(tidybayes)
library(latex2exp)
library(rstan)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

fit_real <- readRDS(here("data/output/trained_sub_real_data_model.rds"))
subset_perfect_data <- readRDS(here("data/output/subset_perfect_data.rds"))

arranged_subset_perfect_data = subset_perfect_data |>
  arrange(tissue,
          cell_line,
          rep,
          ProteinGroup)

n_prot <- 300

n_cell_lines = length(unique(subset_perfect_data$cell_line_name))

n_sample = subset_perfect_data |>
  filter(protein_id == protein_id[1]) |>
  group_by(protein_id) |>
  summarise(n()) |>
  pull(`n()`)

n_tissues <- 2

test_enrichment <- read_csv(here("data/input/enrichment_plot_rep_sum_piv.csv")) |>
  arrange(desc(N)) |>
  separate(Sample, sep ="_", into = c("tissue_name", "sample"), extra="merge") |>
  separate(sample, sep ="_(?=[1-9]$)", into = c("cell_line_name", "rep_name")) |>
  mutate(tissue_name = factor(tissue_name),
         cell_line_name = factor(cell_line_name))
```

```{r explore-posterior}

protein_dictionary <- subset_perfect_data |>
  group_by(ProteinGroup) |>
  summarise(protein_id = protein_id[1]) |>
  mutate(ProteinGroup = factor(ProteinGroup),
         protein_id = unclass(ProteinGroup))

cell_line_dictionary <- subset_perfect_data |>
  group_by(cell_line_name, tissue_name) |>
  summarise(tissue = tissue[1],
            cell_line = cell_line[1])

fit_summary_real <- summary(fit_real)$summary |> 
  as_tibble(rownames = "parameter") |>
  transmute(parameter,
            median = `50%`)

subset_perfect_data_sorted <- arranged_subset_perfect_data |> 
  mutate(ProteinGroup = factor(ProteinGroup),
         protein_id = unclass(ProteinGroup)) |>
  group_by(ProteinGroup) |>
  mutate(sample = rep(1:n_sample))

enrichment_comparison <- subset_perfect_data_sorted |>
          group_by(tissue_name, cell_line_name,
                   rep_name,tissue, cell_line, rep, sample) |> 
          summarise(unique(rep))|>
  inner_join(test_enrichment) |>
  ungroup() |>
  inner_join(fit_real |> 
  spread_draws(enrichment[sample]) |>
  summarise_draws())
  
named_protein_results <- protein_dictionary |> 
  inner_join(fit_real |> 
  spread_draws(chrom_intensity_cell_line_n_min_1[cell_line,protein_id]) |>
  summarise_draws()) |>
  inner_join(cell_line_dictionary)

```

## Posterior Check

Basic sanity check that the model outputs intensity values that match the input intensities.

```{r full_posterior_check}
ggplot( tibble(gen_prot_intensity=c(apply(extract(fit_real,"gen_prot_intensity")$gen_prot_intensity,c(2,3),median))) |>
         mutate(sample = rep(1:n_sample, n_prot),
                protein_id = rep(1:n_prot, each = n_sample)) |>
         inner_join(subset_perfect_data_sorted)) +
  geom_point(aes(x = log(gen_prot_intensity), y = log(prot_intensity)))  +
  labs(x="Predicted Intensity",
       y="Ground Truth Intensity") +
  theme_bw() +
  geom_abline(slope = 1, intercept = 0)
```

## Predicted Enrichment

The relationship is not perfectly linear, as the model predicts greater enrichment the externally calculated enrichment increases for nuclear proteins and decreases for cytoplasmic genes.

```{r nuclear_enrichment}
ggplot(enrichment_comparison) +
  geom_point(aes(x= 1 - median, y=N)) +
  labs(x="Predicted Enrichment",
       y="External Nuclear Enrichment") +
  theme_bw()
```

```{r cytoplasm_enrichment}
ggplot(enrichment_comparison) +
  geom_point(aes(x= 1-median, y=C)) +
  labs(x="Predicted Enrichment",
       y="External Cytoplasmic Enrichment") +
  theme_bw()
```

\newpage

## Predicted protein localisation compared to Kathryn Lilley's hyperlopit results

Although imperfect, there does appear to be a clear divide between proteins expected to be predominately in the cytoplasm and those expected to be in the nucleus.

This is trained on a small subset of the data set (300 genes from 2 tissues).

```{r compare-to-hyperlopit}
protein_localisation_dataset <- read_tsv(here("data/input/annot_hyperlopit.tsv")) |>
  mutate(final.assignment = as.factor(final.assignment))

compare_localisation <- protein_dictionary |> 
  inner_join(fit_real |> 
  spread_draws(cyto_chrom_ratio[protein_id]) |>
  summarise_draws()) |> 
  mutate(Uniprot = ifelse(str_detect(ProteinGroup, ";"),
                               str_extract(ProteinGroup, "[A-Z0-9-]+(?=;)"),
                               as.character(ProteinGroup))) |>
  inner_join(protein_localisation_dataset) |>
  transmute(Uniprot,
            is.cyto = median > 1,
            final.assignment,
            median)

ggplot(compare_localisation) +
  geom_hline(yintercept = 0, colour = "red")+
  geom_boxplot(aes(x = final.assignment, y=log2(median))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90),
        axis.title.x = element_blank()) +
  labs(y = TeX("log $N^{Chr}$ / $N^{Cyt}$")) 
  
```

